FlowFPX: Nimble Tools for Debugging Floating-Point Exceptions

### 2024-08-01 paper.tex
jcon2: all pkg names --> +.jl

\newcommand{\code}[1]{\texttt{#1}}
\newcommand{\FlowFPX}{FlowFPX}
\newcommand{\GPUFPX}{GPU-FPX}
\newcommand{\FloatTracker}{FloatTracker}
\newcommand{\FT}{\FloatTracker}
\newcommand{\Fp}{Floating-point} % hyphen or no?
\newcommand{\fp}{floating-point} % hyphen or no?
\newcommand{\CSTG}{stack graph}
\newcommand{\CPP}{\code{C++}}
\newcommand{\Dendro}{\textsc{Dendro}}
\newcommand{\urlaccess}[2]{\url{#1}}
\newcommand{\Nan}{\code{NaN}}
\newcommand{\NaN}{\Nan}
\newcommand{\Inf}{\code{Inf}}
\newcommand{\zerowidth}[1]{\makebox[0pt][l]{#1}}
\newcommand{\zerocode}[1]{\zerowidth{\code{#1}}}
\newcommand{\genpropkill}{\emph{gen}-\emph{prop}-\emph{kill}}
\newcommand{\bigcheckmark}{\ding{51}}
% \newcommand{\bigxmark}{\ding{53}}
\newcommand{\tblnext}{\(\Rightarrow\)}
% \newcommand{\tblY}{\mbox{\bigcheckmark}}
% \newcommand{\tblN}{\scalebox{0.8}{\bigxmark}}
\newcommand{\tblY}{\checkmark}
\newcommand{\tblN}{\scalebox{1.2}{$\times$}}

@@

This paper introduces \FlowFPX{}, a toolkit for debugging
\fp{} exceptions~(\cref{s:flowfpx})
that has helped improve a variety of applications,
from ocean simulations to heat modes~(\cref{s:casestudies}).
\FlowFPX{} consists of three tools that can work together to debug \fp{} computations:
\begin{enumerate}
  \item
    The centerpiece of \FlowFPX{} is \FT{}, a dynamic analysis tool that
    selectively monitors for exceptions and fuzzes Julia code for vulnerabilities.
    \FT{} is a Julia library that operates on Julia code.
  \item
    To visualize results, \FlowFPX{} uses coalesced stack trace graphs (CSTGs, or
    \CSTG{}s)~\cite{humphreySystematicDebuggingMethods2014}
    to summarize the program contexts that handled exceptional values.
    \CSTG{} is a generic stack trace graph summarization tool.
  \item
    A companion tool \GPUFPX{}~\cite{llsflg-hpdc-2023} provides fine-grained insights for GPU kernels.
    \GPUFPX{} is a dynamic library that programmers of GPU kernels can use.
\end{enumerate}

@@

\FlowFPX{} (FPX: Floating Point eXeption) is a toolkit for tracking down \fp{} exceptions.
The primary tools in the \FlowFPX{} toolkit are \FT{}, which records lifetimes
and enables fuzzing, and \CSTG{}s (more precisely, CSTGs), which visualize
the flow of exceptions.
\GPUFPX{} is a third component of \FlowFPX{} that tracks \fp{} exceptions inside GPU kernels~\cite{llsflg-hpdc-2023}.
This section covers each tool in detail.

\subsection{\FT{}}
\label{s:floattracker}

\FT{} tracks exceptional values in Julia programs across
their \genpropkill{} lifetime and can fuzz code for vulnerabilities by
injecting a \Nan{} or \Inf{} as the result of an operation. \FT{} is
implemented in Julia as a library and is available on JuliaHub:

@@

\FT{} monitors \Nan{} and \Inf{} exceptions by overloading arithmetic
operations and logging key events.
When the input to an operation is exception-free but the output is
exceptional, the operation is a {gen} event~(\cref{fig:lifetime-class}).
When the input and output contain exceptions, the operation is a {prop}
event.
And when the input contains an exception but the output does not,
the operation is a {kill} event.
Put together, the log of all {gen}s, {prop}s, and {kill}s
sheds light of how various exceptions traveled across the program.
The logs also record the call context (stack trace) and the arguments to the operation as a starting point for debugging efforts.

\FT{} writes logs to three files: one for {gen} events, one for {prop}s, and one for {kill}s.
This way, discovering where a \Nan{} came from is a matter of sifting through the {gen} file.
Users can also lower the overhead of logging by turning it off for prop events.

The instrumentation works through custom \fp{} types: \code{TrackedFloat64},
\code{TrackedFloat32}, and \code{TrackedFloat16}.
Developers must opt in to \FT{} by wrapping numbers in a custom type.
From then on, tracking is automatic and extends transitively to all outputs
of tracked operations.
Multiplying a \code{Float64} value with a \code{TrackedFloat64}, for example, yields
a \code{TrackedFloat64} to continue the logging trail.

Crucially, the tracked version of an exception, say a \NaN{}, points
to the original exceptional value.
This enables techniques such as \NaN{}-packing, or otherwise
using the payload in creative ways.
Not all Julia operations currently preserve \NaN{}s
bit-for-bit,\footnote{\url{https://github.com/JuliaLang/julia/issues/48523}}
but \FT{} does its part.

We note that \FT{} is currently not thread-safe; event logs are
collected globally and running \FT{} in a multithreaded environment
carries risks of corrupting the event log stack.
Improving \FT{} to be thread-safe is an area for future work.

@@

In addition to enabling tracking exceptional values, operator
overloading gives \FT{} a powerful way to fuzz code from the inside
out.
Each overloaded function serves as a hook where \FT{} can decide whether to observe
the operation or step in, discard the original result, and return an exception instead.
Injecting faults in a random way~\cite{hamlet1994random}, also known as fuzzing,
is a useful way to discover vulnerabilities in a large codebase.
Demmel et~al. propose essentially the same idea for BLAS and
LAPACK~\cite{ddghlllprr-correctness-2022}.

When fuzzing is enabled, every time \FT{} intercepts a \fp{}
operation, it decides whether or not it should return a \Nan{} instead
of the actual computed value.
This decision is based on several parameters that the developer can
tune to control fuzzing:

@@

\FT{} looks at these parameters in conjunction with stack trace at the
point where the operation was intercepted to make the decision.
When fuzzing is active (\texttt{active == true}), there are \Nan{}s
remaining to be injected (\texttt{n\_inject > 0}), and
\texttt{rand(odds)} returns \texttt{1}, then \FT{} inspects the
current stack trace to see if the point is in the dynamic extend of
the wanted \texttt{functions} or \texttt{libraries}.
When all the conditions are met, the overloaded operator returns
\Nan{} instead of whatever value would have been computed normally.
The value \texttt{n\_inject} is decremented before the next operation
gets intercepted.

Since \FT{} checks if it should inject a \Nan{} at \emph{every} \fp{}
operation it intercepts, means that \Nan{}s can be injected deep
inside of an algorithm instead of just at the inputs.
This can reveal vulnerabilities that otherwise would have been
difficult to catch with input-based fuzzers.

@@

When fuzzing reveals an error, the next step is to craft a regression test to
guide repairs.
\FT{} therefore records the sequence of injections that it makes during a fuzzing
run and enables a replay of any recording after the fact.
Replay runs proceed deterministically so that developers can harden
their code and check that the fixes remove the error.
When running a replay, each intercepted operation consults the reply
file (instead of the stack trace and fuzzing parameters) to decide
whether or not to return a \Nan{} from the computation.

\subsubsection{\FT{} Internals}

\FT{} takes advantage of Julia's operator overloading to track exceptions and
fuzz for vulnerabilities.
For example, below is a variant of \code{+} that is overloaded
for \code{TrackedFloat64}.
It calls the basic \code{+} (or injects a \Nan{}) and checks for exceptions
before returning:
This lets \FT{} intercept all \fp{} operations involving \texttt{TrackedFloat} types.

@@

Every arithmetic operation requires a similar overloading.
Thus, the implementation of \FT{} uses a metaprogramming technique
adapted from \texttt{Sherlogs}~\cite{kMilanklSherlogsJl2021}
to abstract over the common patterns.
For every binary operation, the library creates an overloading similar
to the one for \code{+} above.
Unary operations and others work analogously.
This approach saved thousands of lines of code.
The implementation weighs in at 218 lines and defines 645 overloaded function;
assuming 5 lines of code per function, a handwritten version would require over 3,000 lines.

@@

\FT{} can produce copious amounts of log files which can be challenging to sift through manually.
Coalesced stack trace graphs (CSTGs or \CSTG{}s for short) provide a way to visualize large amounts of stack traces in a compact form~\cite{humphreySystematicDebuggingMethods2014}.
This technique pairs well with \FT{} and makes analyzing the log files easier.

@@

Many programs offload work onto GPUs, which are no less susceptible to \fp{} exceptions than CPUs.
In fact, GPU programs are worse off because they lack exception-handling mechanisms from the CPU world~\cite{llg-soap-2022}.
Since \FT{} instruments Julia programs, it cannot help directly;
however, the companion tool \GPUFPX{} instruments GPU kernels to detect and report
\fp{} exceptions~\cite{llsflg-hpdc-2023}.
Together, \FT{} and \GPUFPX{} provide insights for accelerated programs.

\FT{} and \GPUFPX{} have no formal connection---that is, one does not
need to use \GPUFPX{} with \FT{}, nor vice-versa.
These two tools compliment one another in debugging mixed-paradigm
workloads, but they can be used in isolation.
\GPUFPX{} is a library exclusively for NVIDIA CUDA code and has no
special relation to Julia.
Nevertheless, in cases where Julia code leverages a CUDA kernel,
\GPUFPX and \FT{} can both aid in tracking down exceptional \fp{}
values.

@@

\FlowFPX{} has helped to debug exceptions and fuzz for issues in a variety
of settings, some synthetic and some realistic.
The case studies include a shallow water simulation, the
\texttt{OrdinaryDiffEq} solver, and a Bayesian inference library.

\subsection{ShallowWaters}
\label{s:sw}

\texttt{ShallowWaters} simulates the flow of water over a
seabed~\cite{klowerNumberFormatsError2020,klowerPositsAlternativeFloats2019}.
The library has dozens of parameters that a scientist can experiment with.
One notable parameter is the Courant-Friedrichs-Lewy (CFL) number, which roughly describes the size of the time step to take in running the simulation.
A small CFL number makes the simulation run slowly, but accurately; a large number speeds it up but loses precision
because the system does not get enough time to propagate information.

@@

Running \FT{} on the high-CFL simulation reveals where the simulation drifted
from numbers to \NaN{} exceptions.
The first step in applying \FT{} is to convert relevant floats to tracked
floats, e.g., \code{Float32} to \code{TrackedFloat32}.
For \texttt{ShallowWaters}, this step is easy because the simulation is parameterized by
a \fp{} type for internal use.
Swapping in a tracked type is enough:

@@

\begin{lstlisting}
-([-Inf, -Inf])   FT/TrackedFloat.jl:106
momentum_u!       SW/rhs.jl:246
rhs_nonlinear!    SW/rhs.jl:50
rhs!              SW/rhs.jl:14 [inlined]
time_integration  SW/time_integration.jl:77
run_model         SW/run_model.jl:37
top-level
\end{lstlisting}

@@

\begin{lstlisting}
^([-1.515f31, 2])          FT/TrackedFloat.jl:138
literal_pow()              intfuncs.jl:325
...
materialize(^)             broadcast.jl:860
top-level getproperty(...) examples/sw_nan_tf.jl:14
..
\end{lstlisting}
%
This \Inf{} came from an exponent that overflowed the float type (\code{-1.515e31\string^2}).
\FT{} has shown exactly which numeric values in which operation
caused exceptions to occur.

@@

While the logs for \texttt{ShallowWaters} contain useful information,
there is an overwhelming amount of it.
There are over ten thousand lines in the gen file alone.
Converting these logs to a stack graph gives a quick overview
of the most common paths to exceptional values.

\Cref{fig:sw_nan_cstg} presents the stack graph for \NaN{} gen events
in \texttt{ShallowWaters} with the high CFL number.
Reading bottom-up, every \NaN{} came from calls to the `\code{-}' and `\code{+}' operations.
Calls to `\code{+}' account for most of the \NaN{}s.
These \NaN{}s arose in two different contexts: a small number (30)
occurred within a momentum calculation, while the rest (162)
occurred within a continuity step.
Moving to the top of the graph, it shows that the function \code{run\_model} drove
the entire simulation.
With this overview of the program, a promising next step is to guard against \NaN{}s
in the momentum and continuity functions.

\begin{figure}[t]
  \centering
  \includegraphics[width=0.96\columnwidth]{fig/sw_nan_cstg_clean.pdf}
  \caption{Stack graph for \NaN{} gens in \texttt{ShallowWaters}}
  \label{fig:sw_nan_cstg}
\end{figure}

@@

Graph diffing works well for \CSTG{}s; it shows how flows in the program evolved
from one stage to another.
\Cref{fig:cstg_diff_demo} illustrates one diff in the context of \NaN{} gens for
\texttt{ShallowWaters}:
it compares the first 10\% of gen events to the latter 90\% of gens.
The positive numbers and green lines indicate flows that are new in the latter part,
and the negative numbers and red lines show flows that disappeared in the latter part.
A domain expert might use these clues to find where an instability started in
the first part of the program.
In this case, the function \texttt{momentum\_u!} showed up only in the
beginning of the logs---it may be an effective point to check for \Nan{}s.

@@

\subsection{OrdinaryDiffEq}
\label{s:ode}

With the targeted fuzzing abilities of \FT{}, we set focus on the
\texttt{NBodySimulator} package from the SciML
team.\footnote{\url{https://github.com/SciML/NBodySimulator.jl}}
We only wanted to test the robustness of the \texttt{NBodySimulator} package, so we configured the fuzzer to inject \NaN{}s \emph{only} when inside of a function of that library.
This way, we could avoid injecting in the standard library or other dependencies of \texttt{NBodySimulator} and concentrate on improving this one library.

However, Fuzzing discovered zero bugs and zero exceptional events.
In fact, fuzzing injected \emph{zero} exceptional values whatsoever, even with adjusting the odds to \emph{force} an injection at the first available opportunity.
The reason for this lack of \NaN{} injection, it turns out, is that the nbody
simulation merely creates a problem for the \code{OrdinaryDiffEq}
solver.\footnote{\url{https://github.com/SciML/OrdinaryDiffEq.jl}}
All the \fp{} operations were in another library.

Fuzzing on \code{OrdinaryDiffEq} led to a curious situation.
The library itself reported a \NaN{} and printed a message that it would exit.
However, after printing that message, the program went into an infinite loop.
Using \CSTG{}s to guide our search, we found a \NaN{} kill that manifested repeatedly in the logs.
The stack traces for that kill originated from inside the file \texttt{solve.jl}:

\begin{lstlisting}
<([NaN, 3.0e6])  at FT/TrackedFloat.jl:193
solve!           at ODE/solve.jl:515
...
\end{lstlisting}

@@

Fuzzing with \FT{} revealed two places where Finch needed
protection against user input.
The first was when reading an input mesh.\footnote{\urlaccess{https://github.com/paralab/Finch/issues/16}{2023-06-06}}
A \NaN{} injected in the mesh led to a crash further on:

@@

Before accessing the input, Finch needed to check for \NaN{}s.
The second place was in setting bounds for the
solver.\footnote{\urlaccess{https://github.com/paralab/Finch/issues/17}{2023-06-06}}
Here, a \NaN{} could leave bounds uninitialized, leading to a bounds error.
Additionally, \FT{} and \CSTG{}s have been useful for identifying \NaN{}s that
appear in unstable heat simulations written in Finch.
%% advection2d, high cfl and high T (time) led to NaN during simulation

@@

\subsection{RxInfer}
\label{s:safari}

We discovered an open issue in the \code{RxInfer.jl} library related to \NaN{} detection
and suggested \FT{} to the developers.\footnote{\url{https://github.com/biaslab/RxInfer.jl/issues/116}}
Within a day, they were able to track down the location of an important \NaN{} gen.
This success came with little input from our end; in fact, the application program was
closed-source.
We merely explained how to use \FT{} by wrapping inputs in a tracked type.

@@

Demmel et al.\cite{ddghlllprr-correctness-2022} examine \fp{} exceptional value handling in BLAS and LAPACK, identify several inconsistencies, and propose an API for debugging and adding determinism.
The proposal includes an extension to the \texttt{INFO\_ARRAY} parameter with fields that record \genpropkill{} information.
It also includes fuzzing, which we realize in \FT{}.

@@

Herbie~\cite{panchekhaAutomaticallyImprovingAccuracy2015} automatically rewrites arithmetic expressions to reduce \fp{} error.
This would combine well with \FT{}: first identify the location of a \NaN{}, ask Herbie to find a repair.
The Odyssey\cite{misbackOdysseyInteractiveWorkbench2023} workbench provides an interactive interface to Herbie.

@@

The Julia library \texttt{Sherlogs}~\cite{kMilanklSherlogsJl2021} inspired our use of a custom number type to intercept operations on a number.
In contrast to \FT{} which monitors for exceptional values and logs stack traces at interesting points in their lifetime, \texttt{Sherlogs} tracks and reports the range of values seen over the course of a computation.
This is intended to provide insight into whether or not a library could tolerate a lower-precision \fp{} format.

% TODO: make sure that what I said about what FPSpy can track is true.
FPSpy~\cite{dindaSpyingFloatingPoint2020} is an \texttt{LD\_PRELOAD} shared library that works on unmodified x86 binaries.
It monitors a program during execution for operations that generate an exception, such as division by zero, underflow, and overflow.
By contrast to \FT{}, it does not track prop or kill events.
FPSpy has the advantage of being lightweight enough to run on production code for certain loads.

@@

\FT{} incurs significant overhead on the order of 100x slower than a non-instrumented run of the same program.
It is a debugging tool, not a production tool.
To put this number into context, Valgrind runs with a similar level of slowdown.

>>>>

\newcommand{\code}[1]{\texttt{#1}}
\newcommand{\FlowFPX}{FlowFPX}
\newcommand{\GPUFPX}{GPU-FPX}
\newcommand{\juliapkg}[1]{\texttt{#1.jl}}
\newcommand{\TrackedFloats}{\juliapkg{TrackedFloats}}
\newcommand{\TF}{\TrackedFloats}
\newcommand{\ShallowWaters}{\juliapkg{ShallowWaters}}
\newcommand{\OrdinaryDiffEq}{\juliapkg{OrdinaryDiffEq}}
\newcommand{\RxInfer}{\juliapkg{RxInfer}}
\newcommand{\Sherlogs}{\juliapkg{Sherlogs}}
\newcommand{\NBodySimulator}{\juliapkg{NBodySimulator}}
\newcommand{\Fp}{Floating-point}
\newcommand{\fp}{floating-point}
\newcommand{\CSTG}{stack graph}
\newcommand{\CPP}{\code{C++}}
\newcommand{\Dendro}{\textsc{Dendro}}
\newcommand{\urlaccess}[2]{\url{#1}}
\newcommand{\Nan}{\code{NaN}}
\newcommand{\NaN}{\Nan}
\newcommand{\Inf}{\code{Inf}}
\newcommand{\zerowidth}[1]{\makebox[0pt][l]{#1}}
\newcommand{\zerocode}[1]{\zerowidth{\code{#1}}}
\newcommand{\genpropkill}{\emph{gen}-\emph{prop}-\emph{kill}}
\newcommand{\bigcheckmark}{\ding{51}}
% \newcommand{\bigxmark}{\ding{53}}
\newcommand{\tblnext}{\(\Rightarrow\)}
% \newcommand{\tblY}{\mbox{\bigcheckmark}}
% \newcommand{\tblN}{\scalebox{0.8}{\bigxmark}}
\newcommand{\tblY}{\checkmark}
\newcommand{\tblN}{\scalebox{1.2}{$\times$}}

@@

This paper introduces \FlowFPX{}, a toolkit for debugging
\fp{} exceptions~(\cref{s:flowfpx})
that has helped improve a variety of applications,
from ocean simulations to heat modes~(\cref{s:casestudies}).
\FlowFPX{} consists of three tools that can work together to debug \fp{} computations:
\begin{enumerate}
  \item
    The centerpiece of \FlowFPX{} is \TF{}, a dynamic analysis tool that
    selectively monitors for exceptions and fuzzes Julia code for vulnerabilities.
    \TF{} is a Julia library that operates on Julia code.
  \item
    To visualize results, \FlowFPX{} uses coalesced stack trace graphs (CSTGs, or
    \CSTG{}s)~\cite{humphreySystematicDebuggingMethods2014}
    to summarize the program contexts that handled exceptional values.
    \CSTG{} is a generic stack trace graph summarization tool.
  \item
    A companion tool \GPUFPX{}~\cite{llsflg-hpdc-2023} provides fine-grained insights for GPU kernels.
    \GPUFPX{} is a dynamic library that programmers of GPU kernels can use.
\end{enumerate}

@@

\FlowFPX{} (FPX: Floating Point eXeption) is a toolkit for tracking down \fp{} exceptions.
The primary tools in the \FlowFPX{} toolkit are \TF{}, which records lifetimes
and enables fuzzing, and \CSTG{}s (more precisely, CSTGs), which visualize
the flow of exceptions.
\GPUFPX{} is a third component of \FlowFPX{} that tracks \fp{} exceptions inside GPU kernels~\cite{llsflg-hpdc-2023}.
This section covers each tool in detail.

\subsection{\TF{}}
\label{s:floattracker}

\TF{} tracks exceptional values in Julia programs across
their \genpropkill{} lifetime and can fuzz code for vulnerabilities by
injecting a \Nan{} or \Inf{} as the result of an operation. \TF{} is
implemented in Julia as a library and is available on JuliaHub:

@@

\TF{} monitors \Nan{} and \Inf{} exceptions by overloading arithmetic
operations and logging key events.
When the input to an operation is exception-free but the output is
exceptional, the operation is a {gen} event~(\cref{fig:lifetime-class}).
When the input and output contain exceptions, the operation is a {prop}
event.
And when the input contains an exception but the output does not,
the operation is a {kill} event.
Put together, the log of all {gen}s, {prop}s, and {kill}s
sheds light of how various exceptions traveled across the program.
The logs also record the call context (stack trace) and the arguments to the operation as a starting point for debugging efforts.

\TF{} writes logs to three files: one for {gen} events, one for {prop}s, and one for {kill}s.
This way, discovering where a \Nan{} came from is a matter of sifting through the {gen} file.
Users can also lower the overhead of logging by turning it off for prop events.

The instrumentation works through custom \fp{} types: \code{TrackedFloat64},
\code{TrackedFloat32}, and \code{TrackedFloat16}.
Developers must opt in to \TF{} by wrapping numbers in a custom type.
From then on, tracking is automatic and extends transitively to all outputs
of tracked operations.
Multiplying a \code{Float64} value with a \code{TrackedFloat64}, for example, yields
a \code{TrackedFloat64} to continue the logging trail.

Crucially, the tracked version of an exception, say a \NaN{}, points
to the original exceptional value.
This enables techniques such as \NaN{}-packing, or otherwise
using the payload in creative ways.
Not all Julia operations currently preserve \NaN{}s
bit-for-bit,\footnote{\url{https://github.com/JuliaLang/julia/issues/48523}}
but \TF{} does its part.

We note that \TF{} is currently not thread-safe; event logs are
collected globally and running \TF{} in a multithreaded environment
carries risks of corrupting the event log stack.
Improving \TF{} to be thread-safe is an area for future work.

@@

In addition to enabling tracking exceptional values, operator
overloading gives \TF{} a powerful way to fuzz code from the inside
out.
Each overloaded function serves as a hook where \TF{} can decide whether to observe
the operation or step in, discard the original result, and return an exception instead.
Injecting faults in a random way~\cite{hamlet1994random}, also known as fuzzing,
is a useful way to discover vulnerabilities in a large codebase.
Demmel et~al. propose essentially the same idea for BLAS and
LAPACK~\cite{ddghlllprr-correctness-2022}.

When fuzzing is enabled, every time \TF{} intercepts a \fp{}
operation, it decides whether or not it should return a \Nan{} instead
of the actual computed value.
This decision is based on several parameters that the developer can
tune to control fuzzing:

@@

\TF{} looks at these parameters in conjunction with stack trace at the
point where the operation was intercepted to make the decision.
When fuzzing is active (\texttt{active == true}), there are \Nan{}s
remaining to be injected (\texttt{n\_inject > 0}), and
\texttt{rand(odds)} returns \texttt{1}, then \TF{} inspects the
current stack trace to see if the point is in the dynamic extend of
the wanted \texttt{functions} or \texttt{libraries}.
When all the conditions are met, the overloaded operator returns
\Nan{} instead of whatever value would have been computed normally.
The value \texttt{n\_inject} is decremented before the next operation
gets intercepted.

Since \TF{} checks if it should inject a \Nan{} at \emph{every} \fp{}
operation it intercepts, means that \Nan{}s can be injected deep
inside of an algorithm instead of just at the inputs.
This can reveal vulnerabilities that otherwise would have been
difficult to catch with input-based fuzzers.

@@

When fuzzing reveals an error, the next step is to craft a regression test to
guide repairs.
\TF{} therefore records the sequence of injections that it makes during a fuzzing
run and enables a replay of any recording after the fact.
Replay runs proceed deterministically so that developers can harden
their code and check that the fixes remove the error.
When running a replay, each intercepted operation consults the reply
file (instead of the stack trace and fuzzing parameters) to decide
whether or not to return a \Nan{} from the computation.

\subsubsection{\TF{} Internals}

\TF{} takes advantage of Julia's operator overloading to track exceptions and
fuzz for vulnerabilities.
For example, below is a variant of \code{+} that is overloaded
for \code{TrackedFloat64}.
It calls the basic \code{+} (or injects a \Nan{}) and checks for exceptions
before returning:
This lets \TF{} intercept all \fp{} operations involving \texttt{TrackedFloat} types.

@@

Every arithmetic operation requires a similar overloading.
Thus, the implementation of \TF{} uses a metaprogramming technique
adapted from \texttt{Sherlogs}~\cite{kMilanklSherlogsJl2021}
to abstract over the common patterns.
For every binary operation, the library creates an overloading similar
to the one for \code{+} above.
Unary operations and others work analogously.
This approach saved thousands of lines of code.
The implementation weighs in at 218 lines and defines 645 overloaded function;
assuming 5 lines of code per function, a handwritten version would require over 3,000 lines.

@@

\TF{} can produce copious amounts of log files which can be challenging to sift through manually.
Coalesced stack trace graphs (CSTGs or \CSTG{}s for short) provide a way to visualize large amounts of stack traces in a compact form~\cite{humphreySystematicDebuggingMethods2014}.
This technique pairs well with \TF{} and makes analyzing the log files easier.

@@

Many programs offload work onto GPUs, which are no less susceptible to \fp{} exceptions than CPUs.
In fact, GPU programs are worse off because they lack exception-handling mechanisms from the CPU world~\cite{llg-soap-2022}.
Since \TF{} instruments Julia programs, it cannot help directly;
however, the companion tool \GPUFPX{} instruments GPU kernels to detect and report
\fp{} exceptions~\cite{llsflg-hpdc-2023}.
Together, \TF{} and \GPUFPX{} provide insights for accelerated programs.

\TF{} and \GPUFPX{} have no formal connection---that is, one does not
need to use \GPUFPX{} with \TF{}, nor vice-versa.
These two tools compliment one another in debugging mixed-paradigm
workloads, but they can be used in isolation.
\GPUFPX{} is a library exclusively for NVIDIA CUDA code and has no
special relation to Julia.
Nevertheless, in cases where Julia code leverages a CUDA kernel,
\GPUFPX and \TF{} can both aid in tracking down exceptional \fp{}
values.

@@

\FlowFPX{} has helped to debug exceptions and fuzz for issues in a variety
of settings, some synthetic and some realistic.
The case studies include a shallow water simulation, the
\OrdinaryDiffEq{} solver, and a Bayesian inference library.

\subsection{\ShallowWaters{}}
\label{s:sw}

\ShallowWaters{} simulates the flow of water over a
seabed~\cite{klowerNumberFormatsError2020,klowerPositsAlternativeFloats2019}.
The library has dozens of parameters that a scientist can experiment with.
One notable parameter is the Courant-Friedrichs-Lewy (CFL) number, which roughly describes the size of the time step to take in running the simulation.
A small CFL number makes the simulation run slowly, but accurately; a large number speeds it up but loses precision
because the system does not get enough time to propagate information.

@@

Running \TF{} on the high-CFL simulation reveals where the simulation drifted
from numbers to \NaN{} exceptions.
The first step in applying \TF{} is to convert relevant floats to tracked
floats, e.g., \code{Float32} to \code{TrackedFloat32}.
For \ShallowWaters{}, this step is easy because the simulation is parameterized by
a \fp{} type for internal use.
Swapping in a tracked type is enough:

@@

\begin{lstlisting}
-([-Inf, -Inf])   TF/TrackedFloat.jl:106
momentum_u!       SW/rhs.jl:246
rhs_nonlinear!    SW/rhs.jl:50
rhs!              SW/rhs.jl:14 [inlined]
time_integration  SW/time_integration.jl:77
run_model         SW/run_model.jl:37
top-level
\end{lstlisting}

@@

\begin{lstlisting}
^([-1.515f31, 2])          TF/TrackedFloat.jl:138
literal_pow()              intfuncs.jl:325
...
materialize(^)             broadcast.jl:860
top-level getproperty(...) examples/sw_nan_tf.jl:14
..
\end{lstlisting}
%
This \Inf{} came from an exponent that overflowed the float type (\code{-1.515e31\string^2}).
\TF{} has shown exactly which numeric values in which operation
caused exceptions to occur.

@@

While the logs for \ShallowWaters{} contain useful information,
there is an overwhelming amount of it.
There are over ten thousand lines in the gen file alone.
Converting these logs to a stack graph gives a quick overview
of the most common paths to exceptional values.

\Cref{fig:sw_nan_cstg} presents the stack graph for \NaN{} gen events
in \ShallowWaters{} with the high CFL number.
Reading bottom-up, every \NaN{} came from calls to the `\code{-}' and `\code{+}' operations.
Calls to `\code{+}' account for most of the \NaN{}s.
These \NaN{}s arose in two different contexts: a small number (30)
occurred within a momentum calculation, while the rest (162)
occurred within a continuity step.
Moving to the top of the graph, it shows that the function \code{run\_model} drove
the entire simulation.
With this overview of the program, a promising next step is to guard against \NaN{}s
in the momentum and continuity functions.

\begin{figure}[t]
  \centering
  \includegraphics[width=0.96\columnwidth]{fig/sw_nan_cstg_clean.pdf}
  \caption{Stack graph for \NaN{} gens in \ShallowWaters{}}
  \label{fig:sw_nan_cstg}
\end{figure}

@@

Graph diffing works well for \CSTG{}s; it shows how flows in the program evolved
from one stage to another.
\Cref{fig:cstg_diff_demo} illustrates one diff in the context of \NaN{} gens for
\ShallowWaters{}:
it compares the first 10\% of gen events to the latter 90\% of gens.
The positive numbers and green lines indicate flows that are new in the latter part,
and the negative numbers and red lines show flows that disappeared in the latter part.
A domain expert might use these clues to find where an instability started in
the first part of the program.
In this case, the function \texttt{momentum\_u!} showed up only in the
beginning of the logs---it may be an effective point to check for \Nan{}s.

@@

\subsection{\OrdinaryDiffEq{}}
\label{s:ode}

TODO awk.
With the targeted fuzzing abilities of \TF{}, we set focus on the
\NBodySimulator{} package from the SciML
team.\footnote{\url{https://github.com/SciML/NBodySimulator.jl}}
We only wanted to test the robustness of the \NBodySimulator{} package, so we configured the fuzzer to inject \NaN{}s \emph{only} when inside of a function of that library.
This way, we could avoid injecting in the standard library or other dependencies of \NBodySimulator{} and concentrate on improving this one library.

However, Fuzzing discovered zero bugs and zero exceptional events.
In fact, fuzzing injected \emph{zero} exceptional values whatsoever, even with adjusting the odds to \emph{force} an injection at the first available opportunity.
The reason for this lack of \NaN{} injection, it turns out, is that the nbody
simulation merely creates a problem for the \OrdinaryDiffEq{}
solver.\footnote{\url{https://github.com/SciML/OrdinaryDiffEq.jl}}
All the \fp{} operations were in another library.

Fuzzing on \OrdinaryDiffEq{} led to a curious situation.
The library itself reported a \NaN{} and printed a message that it would exit.
However, after printing that message, the program went into an infinite loop.
Using \CSTG{}s to guide our search, we found a \NaN{} kill that manifested repeatedly in the logs.
The stack traces for that kill originated from inside the file \texttt{solve.jl}:

\begin{lstlisting}
<([NaN, 3.0e6])  at TF/TrackedFloat.jl:193
solve!           at ODE/solve.jl:515
...
\end{lstlisting}

@@

Fuzzing with \TF{} revealed two places where Finch needed
protection against user input.
The first was when reading an input mesh.\footnote{\urlaccess{https://github.com/paralab/Finch/issues/16}{2023-06-06}}
A \NaN{} injected in the mesh led to a crash further on:

@@

Before accessing the input, Finch needed to check for \NaN{}s.
The second place was in setting bounds for the
solver.\footnote{\urlaccess{https://github.com/paralab/Finch/issues/17}{2023-06-06}}
Here, a \NaN{} could leave bounds uninitialized, leading to a bounds error.
Additionally, \TF{} and \CSTG{}s have been useful for identifying \NaN{}s that
appear in unstable heat simulations written in Finch.
%% advection2d, high cfl and high T (time) led to NaN during simulation

@@

\subsection{\RxInfer{}}
\label{s:safari}

We discovered an open issue in the \RxInfer{} library related to \NaN{} detection
and suggested \TF{} to the developers.\footnote{\url{https://github.com/biaslab/RxInfer.jl/issues/116}}
Within a day, they were able to track down the location of an important \NaN{} gen.
This success came with little input from our end; in fact, the application program was
closed-source.
We merely explained how to use \TF{} by wrapping inputs in a tracked type.

@@

Demmel et al.\cite{ddghlllprr-correctness-2022} examine \fp{} exceptional value handling in BLAS and LAPACK, identify several inconsistencies, and propose an API for debugging and adding determinism.
The proposal includes an extension to the \texttt{INFO\_ARRAY} parameter with fields that record \genpropkill{} information.
It also includes fuzzing, which we realize in \TF{}.

@@

Herbie~\cite{panchekhaAutomaticallyImprovingAccuracy2015} automatically rewrites arithmetic expressions to reduce \fp{} error.
This would combine well with \TF{}: first identify the location of a \NaN{}, ask Herbie to find a repair.
The Odyssey\cite{misbackOdysseyInteractiveWorkbench2023} workbench provides an interactive interface to Herbie.

@@

The Julia library \Sherlogs{}~\cite{kMilanklSherlogsJl2021} inspired our use of a custom number type to intercept operations on a number.
In contrast to \TF{} which monitors for exceptional values and logs stack traces at interesting points in their lifetime, \Sherlogs{} tracks and reports the range of values seen over the course of a computation.
This is intended to provide insight into whether or not a library could tolerate a lower-precision \fp{} format.

% TODO: make sure that what I said about what FPSpy can track is true.
FPSpy~\cite{dindaSpyingFloatingPoint2020} is an \texttt{LD\_PRELOAD} shared library that works on unmodified x86 binaries.
It monitors a program during execution for operations that generate an exception, such as division by zero, underflow, and overflow.
By contrast to \TF{}, it does not track prop or kill events.
FPSpy has the advantage of being lightweight enough to run on production code for certain loads.

@@

\TF{} incurs significant overhead on the order of 100x slower than a non-instrumented run of the same program.
It is a debugging tool, not a production tool.
To put this number into context, Valgrind runs with a similar level of slowdown.

In addition to the cost incurred by intercepting \fp{} operations, gathering stack traces is expensive.
We observe a 10x slowdown on \ShallowWaters{} with logging disabled.
We recommend that users of \TF{} make use of the \texttt{maxLogs} and \texttt{exclude\_stacktrace} configuration options to limit the number and kind of logs gathered, and thereby reduce the number of calls to \texttt{stacktrace()}.
Stack traces are essential to decide where to inject a fault when fuzzing,
but we defer them as late as possible to maximize performance.
